```{r}
library(tidyverse)
library(arrow)
library(forecast)
library(tseries)
library(ggplot2)
library(patchwork)
library(knitr)
```
```{r}
INPUT_FILE <- "../../data/processed/merged_features_weekly.parquet"
FREQ <- 52
ALPHA <- 0.05

calc_smape <- function(actual, pred) {
  mean(200 * abs(actual - pred) / (abs(actual) + abs(pred)), na.rm = TRUE)
}
```
```{r}
df <- arrow::read_parquet(INPUT_FILE) %>%
  mutate(date = as.Date(date)) %>%
  arrange(date)

if (!"target_was_missing" %in% names(df)) df$target_was_missing <- 0L

commodities <- unique(df$commodity)
```
```{r}
all_results <- list()
all_models <- list()
all_preds <- list()

for (comm in commodities) {
  sub <- df %>%
    filter(commodity == comm) %>%
    arrange(date)

  train_df <- sub %>%
    filter(split == "train") %>%
    drop_na(target_price)
  test_df <- sub %>%
    filter(split == "test") %>%
    drop_na(target_price)

  if (nrow(train_df) < 30 || nrow(test_df) == 0) next


  n_test <- nrow(test_df)
  train_ts <- ts(train_df$target_price, frequency = FREQ)

  model_auto <- auto.arima(train_ts, seasonal = TRUE, trace = FALSE, ic = "aicc")
  fc_auto <- forecast(model_auto, h = n_test)
  pred_auto <- as.numeric(fc_auto$mean)

  model_111 <- tryCatch(
    Arima(train_ts, order = c(1, 1, 1), include.drift = FALSE),
    error = function(e) NULL
  )
  pred_111 <- if (!is.null(model_111)) as.numeric(forecast(model_111, h = n_test)$mean) else rep(NA, n_test)

  model_sarima <- tryCatch(
    Arima(train_ts, order = c(1, 1, 1), seasonal = list(order = c(1, 1, 1), period = FREQ)),
    error = function(e) NULL
  )
  pred_sarima <- if (!is.null(model_sarima)) as.numeric(forecast(model_sarima, h = n_test)$mean) else rep(NA, n_test)

  pred_df <- tibble(
    date = test_df$date, actual = test_df$target_price,
    was_missing = test_df$target_was_missing,
    auto_arima = pred_auto, arima_111 = pred_111, sarima = pred_sarima
  )

  real_idx <- pred_df$was_missing == 0

  smape_auto <- round(calc_smape(pred_df$actual[real_idx], pred_df$auto_arima[real_idx]), 4)
  smape_111 <- if (!is.null(model_111)) round(calc_smape(pred_df$actual[real_idx], pred_df$arima_111[real_idx]), 4) else NA
  smape_sarima <- if (!is.null(model_sarima)) round(calc_smape(pred_df$actual[real_idx], pred_df$sarima[real_idx]), 4) else NA

  models_fitted <- list(
    "auto.arima" = model_auto,
    "ARIMA(1,1,1)" = model_111,
    "SARIMA(1,1,1)(1,1,1)[52]" = model_sarima
  )

  results <- tibble(
    commodity = comm,
    model = c("auto.arima", "ARIMA(1,1,1)", "SARIMA(1,1,1)(1,1,1)[52]"),
    sMAPE = c(smape_auto, smape_111, smape_sarima)
  ) %>% filter(!is.na(sMAPE))

  all_results[[comm]] <- results
  all_models[[comm]] <- models_fitted
  all_preds[[comm]] <- pred_df
}

results_df <- bind_rows(all_results)
```
### ALL ARIMA MODELS — sMAPE

```{r echo=FALSE}
kable(results_df)
```
```{r}
best_df <- results_df %>%
  group_by(commodity) %>%
  slice_min(sMAPE, n = 1, with_ties = FALSE) %>%
  ungroup()
### BEST MODEL PER COMMODITY (by sMAPE)

```{r echo=FALSE}
kable(best_df)
```
### Residual diagnostics for best models

```{r echo=FALSE}

for (i in seq_len(nrow(best_df))) {
  comm <- best_df$commodity[i]
  best_model <- best_df$model[i]
  model_obj <- all_models[[comm]][[best_model]]

  cat("\n####", comm, "—", best_model, "\n\n")

  resid_clean <- as.numeric(residuals(model_obj))
  resid_clean <- resid_clean[!is.na(resid_clean)]

  lb <- Box.test(resid_clean, lag = 10, type = "Ljung-Box")
  cat(sprintf(
    "  Ljung-Box(10) p-value: %.4f %s\n", lb$p.value,
    ifelse(lb$p.value > ALPHA, "✓", "✗")
  ))

  if (length(resid_clean) <= 5000) {
    nt <- shapiro.test(resid_clean)
    cat(sprintf(
      "  Shapiro-Wilk p-value:  %.4f %s\n", nt$p.value,
      ifelse(nt$p.value > ALPHA, "✓", "✗")
    ))
  }
  cat("\n")
}
```
```{r fig.width=12, fig.height=10}
plots <- list()

for (i in seq_len(nrow(best_df))) {
  comm <- best_df$commodity[i]
  best_model <- best_df$model[i]
  best_smape <- best_df$sMAPE[i]

  sub <- df %>%
    filter(commodity == comm) %>%
    arrange(date)
  train_df <- sub %>%
    filter(split == "train") %>%
    drop_na(target_price)
  test_df <- sub %>%
    filter(split == "test") %>%
    drop_na(target_price)

  cutoff <- max(train_df$date) - 365 * 2
  train_plot <- train_df %>% filter(date >= cutoff)

  pred_df <- all_preds[[comm]]
  pred_col <- switch(best_model,
    "auto.arima" = "auto_arima",
    "ARIMA(1,1,1)" = "arima_111",
    "SARIMA(1,1,1)(1,1,1)[52]" = "sarima"
  )

  test_plot <- pred_df %>% select(date, actual, pred = all_of(pred_col))

  p <- ggplot() +
    geom_line(
      data = train_plot, aes(x = date, y = target_price),
      color = "grey60", linewidth = 0.7
    ) +
    geom_line(
      data = test_plot, aes(x = date, y = actual, color = "Actual"),
      linewidth = 0.9
    ) +
    geom_line(
      data = test_plot, aes(x = date, y = pred, color = "Forecast"),
      linewidth = 0.9, linetype = "dashed"
    ) +
    scale_color_manual(values = c("Actual" = "#2c3e50", "Forecast" = "#e74c3c"), name = NULL) +
    labs(
      title = comm,
      subtitle = sprintf("%s  |  sMAPE = %.2f", best_model, best_smape),
      x = NULL, y = "Price"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 12),
      legend.position = "bottom"
    )

  plots[[i]] <- p
}

if (length(plots) > 0) {
  combined <- wrap_plots(plots, ncol = 2) +
    plot_annotation(
      title = "ARIMA — Best Model Forecast vs Actual",
      theme = theme(plot.title = element_text(face = "bold", size = 14))
    )
  print(combined)
}
```