---
title: "EDA and Stationarity Analysis for 4 Selected Commodities"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(arrow)
library(ggplot2)
library(tseries)
library(forecast)
library(patchwork)
```

## Data Loading

We load the `merged_features_weekly.parquet` dataset which contains the 4 selected commodities.

```{r load_data}
INPUT_FILE <- "../data/processed/merged_features_weekly.parquet"
df <- arrow::read_parquet(INPUT_FILE) %>%
  mutate(date = as.Date(date)) %>%
  arrange(date)

commodities <- unique(df$commodity)
print(commodities)
```

## Raw Time Series Plots

To understand the overall trend and volatility across the different commodities, we visualize their price history.

```{r ts_plots, fig.width=10, fig.height=8}
p_ts <- df %>%
  ggplot(aes(x = date, y = target_price, color = split)) +
  geom_line() +
  facet_wrap(~commodity, scales = "free_y", ncol = 2) +
  theme_minimal() +
  labs(title = "Weekly Prices by Commodity", x = "Date", y = "Price")
print(p_ts)
```

## Stationarity Tests (ADF and KPSS)

We perform the Augmented Dickey-Fuller (ADF) test (null hypothesis: non-stationary) and the KPSS test (null hypothesis: stationary) on the training set for each commodity to determine if differencing is necessary for modeling. We do this for both the raw series and their first differences.

```{r stationarity}
FREQ <- 52

test_results_list <- list()

for (comm in commodities) {
  sub_train <- df %>%
    filter(commodity == comm, split == "train") %>%
    drop_na(target_price) %>%
    arrange(date)
  
  ts_data <- ts(sub_train$target_price, frequency = FREQ)
  diff_ts_data <- diff(ts_data)
  
  adf_raw <- adf.test(ts_data, alternative = "stationary")
  adf_diff <- adf.test(diff_ts_data, alternative = "stationary")
  
  kpss_raw <- kpss.test(ts_data)
  kpss_diff <- kpss.test(diff_ts_data)
  
  test_results_list[[comm]] <- tibble(
    commodity = comm,
    series = c("Raw", "1st Diff"),
    ADF_statistic = c(adf_raw$statistic, adf_diff$statistic),
    ADF_p_value = c(adf_raw$p.value, adf_diff$p.value),
    KPSS_statistic = c(kpss_raw$statistic, kpss_diff$statistic),
    KPSS_p_value = c(kpss_raw$p.value, kpss_diff$p.value)
  )
}

test_results <- bind_rows(test_results_list)
knitr::kable(test_results, digits = 4)
```

## ACF and PACF Analysis

ACF and PACF plots are generated to help identify potential AR and MA terms for ARIMA modeling. 

```{r acf_pacf, fig.width=12, fig.height=12}
for (comm in commodities) {
  sub_train <- df %>%
    filter(commodity == comm, split == "train") %>%
    drop_na(target_price) %>%
    arrange(date)
    
  ts_data <- ts(sub_train$target_price, frequency = FREQ)
  diff_ts_data <- diff(ts_data)
  
  cat("\n### Commodity:", comm, "\n\n")
  
  p1 <- ggAcf(ts_data, main = paste("ACF: Raw -", comm))
  p2 <- ggPacf(ts_data, main = paste("PACF: Raw -", comm))
  p3 <- ggAcf(diff_ts_data, main = paste("ACF: 1st Diff -", comm))
  p4 <- ggPacf(diff_ts_data, main = paste("PACF: 1st Diff -", comm))
  
  print((p1 | p2) / (p3 | p4))
  cat("\n\n")
}
```

## STL Decomposition

We apply STL decomposition to separate the time series into trend, seasonal, and remainder components to better understand underlying patterns.

```{r stl, fig.width=10, fig.height=8}
for (comm in commodities) {
  sub_train <- df %>%
    filter(commodity == comm, split == "train") %>%
    drop_na(target_price) %>%
    arrange(date)
    
  ts_data <- ts(sub_train$target_price, frequency = FREQ)
  
  fit <- stl(ts_data, s.window = "periodic", robust = TRUE)
  
  cat("\n### Decomposition:", comm, "\n\n")
  plot(fit, main = paste("STL Decomposition -", comm))
  cat("\n\n")
}
```
